---
title: "Logging"
author: "Jonathan Callahan"
date: "2018-12-02"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Logging}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Logging in Java and Python

Operational systems, by definition, need to work without human input. Systems
are considered "operational" after they have ben thoroughly tested and shown to
work properly with a variety of input.

However, no software is perfect and no real-world system operates with 100% 
availability or 100% consistent input. Things occasionally go wrong -- perhaps
intermittently. In a situation with occasional failures it is vitally important
to have good logging information for forensic analyis.

Other languages used in operational settings support logging at different levels
of information density. The code to write out logs at these different levels
looks very similar in java and python:

**java**
```
import org.apache.commons.logging.Log;
import org.apache.commons.logging.LogFactory;
import org.apache.commons.logging.impl.Jdk14Logger;
```
... ugly setup of log files ...
```
log.error("ERROR level message");
log.warn("WARN level message");
log.info("INFO level message");
log.debug("DEBUG level message");
log.trace("TRACE level message");
```

**python**
```
import logging
 
logging.basicConfig(filename="my_INFO.log",
                    level=logging.INFO)

logging.error("ERROR level message")
logging.warning("WARNING level message")
logging.info("INFO level message")
logging.debug("DEBUG level message")
logging.trace("TRACE level message")
```

# Logging in R

The **MazamaCoreUtils** package provides functions so that R logging can look
very similar:

```
library(MazamaCoreUtils)
 
logger.setup(infoLog="my_INFO.log")

logger.error("ERROR level message")
logger.warn("WARN level message")
logger.info("INFO level message")
logger.debug("DEBUG level message")
logger.trace("TRACE level message")
```

This functionality is built on top of the excellent
[futile.logger](https://github.com/zatonovo/futile.logger) package.

## Log Levels

When used properly, logging at different levels lets you to create log files 
that help you quickly navigate to the specific part of your code base that 
generated an error.

Different log levels also allow problems to be dealt with by staff with different
skill levels. Core developers don't need to be dragged in whenever anything goes
wrong. If a real-time web site is non-responsive, non-technical staff can be 
assigned to look at logs to assess the situation.

 * If the `ERROR` or `WARN` logs report that no data was obtained from an upstream 
source, the problem is not with the R code.
 * If the `INFO` log reports that incoming data was properly obtained but that
none of it passed QC thresholds, the problem is not with the R code.
 * If, however, these logs show everything running smoothly until some ugly R
error message, then it is time to call in the core developers.

Here are our best practices for how to use the different log levels:

### ERROR

`ERROR` log statements should only be generated immediately before a `stop()` is
issued. The `ERROR.log` file should only have entries when R code failed to
complete.

### WARN

`WARN` log statements imply that something unexpected happend but that R code is
able to complete, perhaps with unpredicatble results. An appropriate use would
be after testing the response of an internet request. R code may be able to
trap and handle a failed request but should comment on the failure in the logs.

### INFO

`INFO` log statements are useful for a general overview of how your R code is
proceeding, reflecting top level code structure and perhaps ending with a final:

```
  logger.info("Successfully completed ...")
  logger.info("\n=============== THE END ===============\n")
```

### DEBUG

`DEBUG` log statements are best used to map out the path taken through the
code. We recommend including `DEBUG` level statements inside of top level loops
and beginning each function with the following:

```
logger.debug("----- FUNCTION_NAME() -----")
```

### TRACE

`TRACE` log statements are for the gory details of any part of your code where
you expect or are experiencing errors. These should contain information about
program state and are intended only for developers.

# A Logging Example

The following minimal example demonstrates the use of logging statements and
their output. 

```{r loggingExample}
library(MazamaCoreUtils)

# Set up logging 
tmpDir <- tempdir()
logger.setup(
  errorLog = file.path(tmpDir,"ERROR.log"),
  infoLog = file.path(tmpDir,"INFO.log"),
  traceLog = file.path(tmpDir,"TRACE.log")
)

# Begin main program
logger.info("Begin loggingExample ...")

# Try something that might fail
outputSums <- list()
for ( column in names(iris) ) {
  
  logger.trace("Working on: %s", column)
  
  result <- try({
    outputSums[[column]] <- 
      paste0("Sum of ", column, " = ", sum(iris[[column]]))
  }, silent = FALSE)
  
  if ( "try-error" %in% class(result) ) {
    logger.warn("Oops: %s", geterrmessage())
  }
  
}

# Successful completion
logger.error("NO ERRORS")
logger.info("Successfully completed loggingExample")
logger.info("\n=============== THE END ===============\n")

# Check the output
str(outputSums)
```

Here is what is generated in the log files:

## ERROR.log

```{r errorLog, echo = FALSE}
cat(paste(readLines(file.path(tmpDir,"ERROR.log")),collapse='\n'))
```

## INFO.log

```{r infoLog, echo = FALSE}
cat(paste(readLines(file.path(tmpDir,"INFO.log")),collapse='\n'))
```

## TRACE.log

```{r traceLog, echo = FALSE}
cat(paste(readLines(file.path(tmpDir,"TRACE.log")),collapse='\n'))
```

# Extra Functionality

Two more utility functions need to be mentioned

### `logger.setLevel()`

This function sets the log level used in the console. By default this level is
set to `FATAL` which basically means "only when R crashes". If you are debugging
code interactively you may wish to see log messages as the are generated by
entering the following command:

```
logger.setLevel(TRACE)
```

### `initializeLogging()`

This function is a wrapper for our typical usage. It accepts a `logDir`
parameter and looks for `TRACE`, `DEBUG` `INFO` and `ERROR` logs in that 
directory. If they are found, they are renamed as 
`paste0(logLevel,".log.",timestamp)`. After renaming, new log files are setup.

This is particularly useful for operational code run inside docker containers
which may be automatically rebooted.



